---
title: "Homework 16"
author: "Stanley Chan"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above. For any included plots, make sure they are clearly labeled. You are free to use any plot type that you feel best communicates the results of your analysis.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

### Install Biostrings
```{r,bquiet=TRUE, message=FALSE, warning=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("Biostrings")
```
## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(dplyr)
library(readr)
library(Biostrings)
``` 

```{r}
options(scipen = 999)
```


## Background
A patient has been admitted to the hospital with a severe and rapidly progressing bacterial infection of unknown origin. Standard clinical cultures have been inconclusive. As part of the diagnostic workup, RNA was extracted directly from the patient's tissue sample and submitted for RNA sequencing (RNA-seq). From this sequencing data, you have successfully assembled the complete coding sequence of a gene expressed by the infecting bacterium.
You will use BLAST, a tool used in real clinical and research microbiology labs, to compare this unknown sequence against all known biological sequences in the NCBI database.

**1. Read the "gene of interest" sequence from the `goi_01.txt` file into R using the Biostrings package. Store this in an object called `query_seq`.What type sequence is this? (DNA, RNA or Protien) **

```{r}
query_seq <- readDNAStringSet("data/goi_01.fasta")
```

Based on the nucleotides in goi_01.fasta, this is a DNA sequence.
```{r}
query_seq
```


**2.Perform a blastx search using `goi_01.txt` as the query and the "nr_cluster_seq" database. Download the cluster table results in CSV format, change the name to "goi_01_blastx.csv" and store in the homework/data folder. Read the csv file into R and clean the column names. Save the cleaned data frame as goi_01_blastx.**
```{r}
goi_01_blastx <- read_csv("data/goi_01_blastx.csv") %>% 
  clean_names()
```

```{r}
goi_01_blastx
```

**3.Display the top 10 blastx hits, based on e_value, for this blastx search? (Include the Scientific Name, Cluster Anscestor, Cluster Representative Sequence, Max Score, Query Cover, E value and Percent Identity)**
```{r}
blastx_hot_10 <- goi_01_blastx %>% 
  select(scientific_name, cluster_ancestor, representative_sequence, max_score, query_cover, e_value, per_ident) %>% 
  arrange(desc(e_value)) %>% 
  slice_head(n=10)
blastx_hot_10
```
 

**4. Create a plot that shows the different in percent identity for each Scientific Name of the top 10 blastx hits. Make sure to label your axes and give your plot a title.**

For clarification on the new mutated columns, I was having a hard time splitting some of the columns because they shared the same scientific name and started off with columns exceeding 100%. I asked Copilot what the issue was and they brought that scientific name issue and suggested that I apply a rank to each of the scientific names of the top 10 blastx hits to make them unique. I've used and annotated their rank and label codes here.
```{r}
blastx_hot_10 %>%
  mutate(rank = row_number(), # Apply a rank based on e_value as that was how I arranged the data set before
         label = paste0(rank, ". ", scientific_name)) %>%  # Concatenate ranking and scientific name
  ggplot(aes(x=label, y=per_ident, fill=scientific_name))+
  geom_col()+
  coord_flip()+
  labs(title = "Percent Identity of Top 10 Blastx Hits", x = "Scientific Name", y = "Percent Identity (%)")+
  theme(legend.position="none")
```

**5 Perform a blastn search using `goi_01.txt` as the query and the "core_nt" database. Download the alignment description in CSV format, change the name to "goi_01_blastn.csv" in the homework folder. Read the csv file into R and clean the column names. Save the cleaned data frame as goi_01_blastn.**
```{r}
goi_01_blastn <- read_csv("data/goi_01_blastn.csv") %>% 
  clean_names()
```

```{r}
goi_01_blastn
```

**6. What was the top blastn hit, based on e_value, for this blastn search? (Include the Cluster Ancestor, Cluster Representative Sequence, Max Score, Query Cover, E value, Percent Identity and Accession Link)**

I didn't see a cluster ancestor or cluster representative sequence on the chart so I have included what I can about the top blastn hit. Looking at the data, all the rows seem to have an e_value of 0 and a max score of 688. For this assignment, I used the blastn hit with the accession term "CP045155.1".
```{r}
goi_01_blastn %>% 
  select(scientific_name, max_score, query_cover, e_value, per_ident, accession) %>% 
  arrange(desc(e_value)) 
```

**7. Based on the blastx and blastn results, what do you think this gene is and what bacteria do you think it came from?**

From the results, this gene is likely from the pMT1 plasmid from the bacteria Yersinia pestis. This gene likely codes for the hypothetical protein YpF1991016_2825.

**8. For the top blastn hit, click on the accession link for this bacteria and download the coding sequences as a FASTA Protein file. Read the fasta file into R using the Biostrings package and store it in an object called `goi_blastn_hit`.**

For reference, top blastn hit in my table is labelled as "Yersinia pestis strain SCPM-O-B-5935 (I-1996) plasmid pMT, complete sequence" on the website.
```{r}
goi_blastn_hit <- readDNAStringSet("data/sequence.fasta")
```

```{r}
goi_blastn_hit
```

**9. From the same organism, identify 5 other annotated genes? Do these genes suggest about the biology or virulence of this bacterium?** *Annotated genes can be found in the "CDS" section of the GenBank file. You can also find genes by going to summary page for the organism and looking at the "View annotated genes" section.*

```{r}
annotated_genes <- read_tsv("data/ncbi_dataset.tsv")
```

These genes are related to both biology and virulence of this bacterium as they would need DNA-replicating genes for survival and reproduction to spread disease.
```{r}
annotated_genes
```


**10. Now that you have an idea of this bacteria's identity, what illness/disease do you think the patient has? What is the historical or public health significance of this disease? Based on the gene you identified, what role might this gene play in pathogenicity?**

I think this patient has plague. Yersinia pestis is historical known as the bacteria that caused the Black Death in Europe. The gene likely plays a role in the reproducibility in the host as DNA replicating genes would help it multiply quickly to further the host's infection.


## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! 